---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# scMetaTraj

`scMetaTraj` is an R package for metabolism-centered trajectory analysis in
single-cell data. The package scores curated metabolic modules, embeds cells in
a metabolic feature space, identifies metabolic subclusters, and infers
metabolic pseudotime (mPT) from graph distances in that space.

## Installation

The development version can be installed from GitHub:

```{r install-github, eval=FALSE}
# install.packages("pak")
pak::pak("Dr-xyGreg/scMetaTraj")
```

After CRAN release, install with:

```{r install-cran, eval=FALSE}
install.packages("scMetaTraj")
```

## Core workflow

The package is organized around five main steps:

1. score pathway-level metabolic activity with `scMetaTraj_score()`
2. embed cells with `scMetaTraj_embed()`
3. identify metabolic subclusters with `scMetaTraj_cluster()`
4. infer metabolic pseudotime with `scMetaTraj_infer()`
5. quantify dynamic module changes with `scMetaTraj_trend()` and related
   plotting helpers

## Minimal example

The example below uses a small simulated expression matrix and toy metabolic
gene sets so that it can run without private data files.

```{r example}
library(scMetaTraj)

set.seed(123)

expr <- matrix(
  rexp(12 * 80, rate = 1),
  nrow = 12,
  ncol = 80,
  dimnames = list(
    c(
      "HK1", "PFKP", "LDHA", "GPI", "CS", "ACO2",
      "NDUFA1", "COX4I1", "G6PD", "PGD", "ACLY", "FASN"
    ),
    paste0("Cell", seq_len(80))
  )
)

gene_sets <- list(
  Glycolysis = c("HK1", "PFKP", "LDHA", "GPI"),
  TCA = c("CS", "ACO2"),
  OXPHOS = c("NDUFA1", "COX4I1"),
  PPP = c("G6PD", "PGD"),
  Lipid = c("ACLY", "FASN")
)

scores <- scMetaTraj_score(expr, gene_sets, min_genes = 2, scale = FALSE)
emb_pca <- scMetaTraj_embed(scores, method = "PCA", n_pcs = 4)
clusters <- scMetaTraj_cluster(emb_pca, k = 10, method = "louvain")
traj <- scMetaTraj_infer(emb_pca, k = 10, root_mode = "pc1_min")

trend_df <- scMetaTraj_trend(
  scores = scores[, "Glycolysis"],
  mPT = traj$mPT,
  n_bins = 20
)

head(trend_df)
table(clusters)
```

## Package scope

`scMetaTraj` is intended to complement transcriptome-centered trajectory tools.
It is most useful when the primary biological question concerns continuous
metabolic remodeling rather than transcriptomic lineage structure alone.

## Documentation

- The main workflow vignette is in `vignettes/scMetaTraj-workflow.Rmd`.
- Function-level help is available with `?scMetaTraj_score`,
  `?scMetaTraj_infer`, and related topics.
